Estimating duration in the presence of misclassification

### 2024-03-29 cis-perfect-testing.tex
proof

In \cref{perf-test:sec:problem}, I describe the challenges analysing the data presents.
This motivates using a survival analysis framework which I introduce in \cref{perf-test:sec:surv-analysis}, before deriving the statistical model for the CIS setting in \cref{perf-test:sec:model}.
In \cref{perf-test:sec:parameters-priors}, I consider how to parameterize the survival function and choose its priors, including vague and smoothing priors as well as a prior constructed to include the information from \cref{E-ATACCC}.
I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they successfully recover $S(t)$.
Finally, I discuss the results in \c{perf-test:sec:discussion}.

@@

Another important concepts typical of survival data is truncation, which refers to ...?
Sometimes, only intervals satisfying some property are recorded.
In this case, data are \emph{truncated}.
\emph{Left truncation} is when only events lasting longer than a minimum amount of time are observed.
\emph{Right truncation} is when only events lasting shorter than a maximum amount of time are observed.
Without correcting for truncation, inference on the distribution of a duration will be biased.

@@

I will use these regions to form the data likelihood, conditional on $\ntot$.
The definition of the admissible region relies on the test results.
However the likelihood should be conditional only on $\vec{\theta}$, $\ntot$, and the test schedules.
Therefore, this is not strictly correct.
This issue can be rectified be considering the inadmissible region as many subregions, which leads to the same result (sketched in \cref{perf-test:sec:alt-likelihood}).
A full formalisation of the derivation in \cref{perf-test:sec:alt-likelihood} would require much more complex notation for the same result; therefore, I prefer the presentation to follow.

@@

When constructing the prior, two aspects need consideration.
Firstly, the model structure from \cref{E-ATACCC} leads to the posterior distribution on the hazards having a positive correlation between $\lambda_t$ and $\lambda_{t'}$, especially when $t$ and $t'$ have a small difference.
The prior used in this analysis should preserve this correlation.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated because it extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
%Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior to include this prior information into the CIS analysis two steps.

>>>>

In \cref{perf-test:sec:problem}, I describe the challenges analysing the data presents.
This motivates using a survival analysis framework which I introduce in \cref{perf-test:sec:surv-analysis}, before deriving the statistical model for the CIS setting in \cref{perf-test:sec:model}.
In \cref{perf-test:sec:parameters-priors}, I consider how to parameterize the survival function and choose its priors, including vague and smoothing priors as well as a prior constructed to include the information from \cref{E-ATACCC}.
I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they successfully recover $S(t)$.
Finally, I discuss the results in \cref{perf-test:sec:discussion}.

@@

Sometimes, only intervals satisfying some property are recorded.
In this case, data are \emph{truncated}.
\emph{Left truncation} is when only events lasting longer than a minimum amount of time are observed.
\emph{Right truncation} is when only events lasting shorter than a maximum amount of time are observed.
Without correcting for truncation, inference on the distribution of a duration will be biased.

@@

I will use these regions to form the data likelihood, conditional on $\ntot$.
The definition of the admissible region relies on the test results.
However the likelihood should be conditional only on $\vec{\theta}$, $\ntot$, and the test schedules.
Therefore, this is not strictly correct.
This issue can be rectified by considering the inadmissible region as many subregions, which leads to the same result (sketched in \cref{perf-test:sec:alt-likelihood}).
A full formalisation of the derivation in \cref{perf-test:sec:alt-likelihood} would require much more complex notation for the same result; therefore, I prefer the presentation to follow.

@@

When constructing the prior, two aspects need consideration.
Firstly, the model structure from \cref{E-ATACCC} leads to the posterior distribution on the hazards having a positive correlation between $\lambda_t$ and $\lambda_{t'}$, especially when $t$ and $t'$ have a small difference.
The prior used in this analysis should preserve this correlation.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated because it extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
%Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior to include this prior information into the CIS analysis in two steps.

### 2024-03-29 incidence-prevalence.tex
proof

\emph{Duration}, or \emph{time-to-event}, distributions are expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of positivity.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variable; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.

@@

Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of a corresponding stochastic model.
When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
The approximation works especially well if used to make infernce based on noisy observations of the system, when the observation noise dominates is much larger than the infection process noise; I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.

@@

The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
The extension adds adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
The ODEs are now:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t) \\
\lambda(t) &= \beta \sum_{j=1}^n i_j(t)
\end{align}
where $n$ is the number of infectious compartments.
The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterization is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.

@@

The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i}_j(t)$ where $\vec{i}_j$ is a vector such that $i_{j,a}$ is the proportion of stratum $a$ in the $j$th infectious state.
Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Following the same argument as in \cref{SEIR:sec:SIR} finds that the infectious individual makes contact with susceptible individuals in stratum $a$ at a mean rate of $\kappa_{aa'} s_a(t)$ per day.
Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
Therefore, the mean rate of infections in strata $a$ per day caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, and dividing by the size of strata $a$, $N_a$, gives $\kappa_{aa'} q_{aa'} s_a(t) \frac{1}{N_a} N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} \frac{1}{N_a} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
This gives the total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.

@@

Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocite{birrellBayesian,jacksonEffects}.
Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
In particular, the contacts are \emph{associative} (contacts are more likely between individuals of similar ages) and younger age groups have more contacts; this will tend to concentrate the epidemic in these younger age groups~\autocite[67]{keelingModeling}.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{inc-prev/contact_matrices}}
  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}; (B) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12. A clear diagonal, showing associative mixing is visible, and children have more contacts. Parallel diagonals off the main diagonal show mixing between parents and their children. Figure adapted from \textcite{klepacContacts}, licensed under \href{https://creativecommons.org/licenses/by-nc-nd/4.0/}{CC BY-NC-ND 4.0.}}
  \label{SEIR:fig:age-contacts}
\end{figure}

@@

On day $t$ (for $t = 1, \dots, T$), a prevalence survey such as the CIS samples $n_t$ individuals at random from the population of interest and tests if they are prevalent.
Therefore, I model the number of positive tests, $Y_t$, as $Y_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
These quantities are fixed by the study design and population choice, and hence I assume independent of independent of $P_t$ and $Z_t$.

>>>>

\emph{Duration}, or \emph{time-to-event}, distributions are expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of detectability.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variable; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.

@@

Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of a corresponding stochastic model.
When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
The approximation works especially well if used to make inference based on noisy observations of the system, when the observation noise is much larger than the infection process noise; I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.

@@

The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
The extension adds multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
The ODEs are now:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t) \\
\lambda(t) &= \beta \sum_{j=1}^n i_j(t)
\end{align}
where $n$ is the number of infectious compartments.
The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterization is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.

@@

The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i}_j(t)$ where $\vec{i}_j$ is a vector such that $i_{j,a}$ is the proportion of stratum $a$ in the $j$th infectious state.
Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Following the same argument as in \cref{SEIR:sec:SIR} finds that the infectious individual makes contact with susceptible individuals in stratum $a$ at a mean rate of $\kappa_{aa'} s_a(t)$ per day.
Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
Therefore, the mean rate of infections in strata $a$ per day caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, and dividing by the size of strata $a$, $N_a$, gives $\kappa_{aa'} q_{aa'} s_a(t) \frac{1}{N_a} N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} \frac{1}{N_a} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
This gives the total force of infection on strata $a$ from all strata is $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.

@@

Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocite{birrellBayesian,jacksonEffects}.
Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
In particular, the contacts are \emph{associative} (contacts are more likely between individuals of similar ages) and younger age groups have more contacts; this will tend to concentrate the epidemic in these younger age groups~\autocite[67]{keelingModeling}.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{inc-prev/contact_matrices}}
  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12; (B) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}. A clear diagonal, showing associative mixing is visible, and children have more contacts. Parallel diagonals off the main diagonal show mixing between parents and their children. Figure adapted from \textcite{klepacContacts}, licensed under \href{https://creativecommons.org/licenses/by-nc-nd/4.0/}{CC BY-NC-ND 4.0.}}
  \label{SEIR:fig:age-contacts}
\end{figure}

@@

On day $t$ (for $t = 1, \dots, T$), a prevalence survey such as the CIS samples $n_t$ individuals at random from the population of interest and tests if they are prevalent.
Therefore, I model the number of positive tests, $Y_t$, as $Y_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
These quantities are fixed by the study design and population choice, and hence I assume independent of $P_t$ and $Z_t$.
